Abstractions for Multi-Sorted Substitutions

### 2023-08-30 paper/paper.tex
Update.

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      Appendix TODO explains the workaround.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

@@

  Now we are finally ready to define the actual kit instances.
  The variable kit definition is straightforward:
  \LibCode\KKitVar

@@

  \section{Map Composition}
  An operation we did not cover in the previous section is the
  composition of substitutions.
  This operation and its properties are relevant in numerous places, e.g.\
  TODO, which are also relevant for System F.

  As we defined substitution in terms of renamings, we now have to
  consider all four compositions between renamings and substitutions,
  and their properties. The problem is again that they are dependent
  on each other, i.e.\ composition between two substitutions relies on
  the compositions between renamings and substitutions, which rely on
  compositions between two renamings.

@@

  In contrast we define a structure similar to Kits, which allows us
  to abstract over all four compositions and use the same trick as
  before to eliminate the dependencies.
  This has the benefit that a user of our library has to write less
  boilerplate and also that we gain more expressiveness, which helps
  us later to unify type preservation of all substitution and renaming
  operations into a single lemma.

@@

  \section{Future Work}

>>>>

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      Appendix TODO explains the workaround.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

@@

  Finally, we define the actual kit instances.
  The variable kit definition is straightforward:
  \LibCode\KKitVar

@@

  \section{Map Composition}
  In this section we extend our framework with the composition of
  arbitrary maps.

  Composition of substitutions is important for most language
  formalizations, as it allows to reason about the meaning of applying
  multiple substitutions to a term. This appears for example in the proof of
  substitution-preserves-typing for System F, where in the
  case of type-application \ACode{(e \ACon{∙} t) \ADef{⋯} σ} we need to reason about
  \ACode{(e \ADef{⋯} \ADef{⦅} t \ADef{⦆}) \ADef{⋯} σ}, which is equivalent to
  \ACode{e \ADef{⋯} (\ADef{⦅} t \ADef{⦆} \ADef{·ₖ} σ)},
  where \ACode{\_\ADef{·ₖ}\_} is backward composition.

  As we defined substitution in terms of renamings, we need to
  consider all four compositions between renamings and substitutions.
  While the composition operations are defined independently of each other,
  the \ACode{\AField{⋯-assoc}} lemma for two substitutions, depends on the
  \ACode{\AField{⋯-assoc}} lemmas between a renaming and a substitution, which in
  turn depend on the \ACode{\AField{⋯-assoc}} lemma between two renamings.

@@

  In contrast we define structures similar to \ACode{\ACon{Kit}} and
  \ACode{\ACon{Traversal}}, which allows us to abstract over all four
  compositions and use the same trick as before to eliminate the
  dependencies.
  This has the benefit that a user of our library has to write less
  boilerplate, but more importantly it allows to prove lemmas about the
  composition of arbitrary maps, which enables us to later unify type
  preservation of all substitution and renaming operations into a single
  lemma.

@@

  \section{Future Work}
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}.
  \end{itemize}
